---
title: "Merging samples"
subtitle: "MANAAB + Franjic // SoupX/DoubletFinder/DropletQC Pipeline"
author: "Marcos Nascimento"
date: "`r Sys.Date()`"
output: html_notebook
---
```{r Setup}
# Loading necessary libraries
library(tidyverse)
library(Seurat)
library(viridis)
library(patchwork)
library(DropletUtils)
library(future)
library(tibble)
library(ComplexHeatmap)

# Additional color scales
library(MetBrewer)
library(NatParksPalettes)

# Parallel processing setup
plan("multicore", workers = 80)
options(future.globals.maxSize = 30 * 1024^3) # 30 GB

# Custom themes and scales for plots
mytheme <- theme_minimal() + 
  theme(axis.line = element_line(),
        axis.ticks = element_line(),
        text = element_text(family = "Helvetica"))

simple <- NoAxes() + NoLegend()
mysc <- scale_color_viridis(option = "A")
region.pal <- c("#5EBFA2", "#F69663", "#731DD8",  "#FB7C7E")

# List of sex-specific genes
sex.genes <- c("TTTY14", "NLGN4Y", "USP9Y", "UTY", "XIST", "RPS4X", "TMSB4X", "TSIX")

```

#Creating seurat objects from count matrices 
Count matrices are downloaded from GEO and saved in a folder named "matrices"
```{r}
samples <- c("CGE", "MGE", "LGE",  "EC Stream", "dEC", "H71", "H31", "H37", "hsb231", "hsb237", "hsb628", "H48", "H39", "H46","H29", "H33")

srt_objects <- list()

for (s in samples){
  cat(paste0("Importing sample ", s, "\n"))
  matrix <- ReadMtx(mtx = paste0("matrices/", s, "_counts.mtx"), 
                    features = paste0("matrices/", s, "_genes.tsv"), feature.column = 1, 
                    cells = paste0("matrices/", s, "_barcodes.tsv"))
  metadata <- read.csv(paste0("matrices/", s, "_metadata.csv"), row.names = 1)
  srt_objects[[s]] <- CreateSeuratObject(counts = matrix, meta.data = metadata)
}
```

Merging samples in a single Seurat object
```{r}
all.exp = merge(srt_objects[[1]], srt_objects[-1])
```

## log Normalization
```{r}
all.exp_raw@active.assay = "RNA"
all.exp_raw = all.exp_raw %>% 
              NormalizeData(assay = "RNA", 
                            verbose = F) %>% 
              FindVariableFeatures() %>% 
              ScaleData()
  
VariableFeatures(all.exp_raw@assays$RNA) = all.exp_raw@assays$RNA@var.features[!(all.exp_raw@assays$RNA@var.features %in% sex.genes)]


all.exp_raw = all.exp_raw %>% RunPCA(npcs = 50)

all.exp_raw = all.exp_raw %>% 
                  RunUMAP(dims = 1:22) %>% 
                  FindNeighbors(dims = 1:22) %>%
                  FindClusters(resolution = c(0.8, seq(0.5, 2, 0.5)))

```


# Overview FeaturePlots
```{r}
FeaturePlot(all.exp, c("GFAP", "HOPX", "TOP2A", "EOMES", "DCX", "TBR1", "SLC17A7", "GAD2", "LHX6", "NR2F2", "PROX1", "CALB2", "LAMP5", "RELN", "VIP", "NPY", "SST", "OLIG2", "SOX10",  "MBP", "FOXJ1", "PECAM1", "PDGFRB", "ADAM28"), order = F, raster = F, ncol = 6) &
  simple &
  mysc & 
  coord_fixed()

ggsave("all.exp_featplots.png", width = 10, height = 6, scale = 2)
```

# Overview Plots
Main Cell Types
```{r}
DimPlot(all.exp, raster = F, label = T, repel = T, group.by = "all.exp_type", shuffle = T) +
  coord_fixed() +
  simple +
  labs(title = "Cell Types")
```

Unsupervised Clutering
```{r}
DimPlot(all.exp, raster = F, label = T, group.by = "RNA_snn_res.1", shuffle = T) + 
  coord_fixed() + 
  simple+
  scale_color_manual(values = met.brewer("VanGogh2", n = 50, override.order = T))+ 
  labs(title = "Unsupervised Clustering")
```

Donor Age
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "age", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_viridis_d(option = "H", name = "Donor Age")+ 
  labs(title = NULL)
```

Donor Age Group
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "age_group", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_manual(values=met.brewer("Hokusai3"), name = "Donor Age Group")+ 
  labs(title = NULL)
```

Sample ID
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "sample", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(values=met.brewer("Juarez", 16), name = "Sample")+ 
  labs(title = NULL)
```

Sample Origin
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "region", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(name = "Sample Origin", values = region.pal) + 
  labs(title = NULL)
```

EC Stream Cells
```{r}
DimPlot(all.exp, group.by = "stream_highlight", order = T, raster = F) + 
  scale_color_manual(values = (region.pal)[3], na.value = "grey85", labels = c("EC Stream (14d)", "Other cells")) +
  coord_fixed() +
  NoAxes() + 
  labs(title = "EC Stream Cells")
```

Nuclear Fraction
```{r}
FeaturePlot(all.exp, "nuclear_fraction", order = T, raster = F) +
  scale_color_viridis(limits = c(0, 1), name = "Nuclear Fraction") +
  coord_fixed() +
  NoAxes()+
  labs(title = NULL)
```

# Saving all.exp
```{r}
Idents(all.exp) = "all.exp_type"
all.exp <- BuildClusterTree(all.exp, assay = "RNA", reorder = T, features = all.exp@assays$RNA@var.features)

saveRDS(all.exp, file = "all.exp.rds", compress = F)
```

# Subsetting interneurons
```{r}
inter.exp.cells = all.exp@meta.data %>% filter(all.exp_type == "Cortical Interneurons") %>% rownames()
DimPlot(all.exp, label = T, cells.highlight = inter.exp.cells, raster = F) + simple + coord_fixed()

inter.exp = subset(all.exp, cells = inter.exp.cells)
```

## logNormalization
```{r}
inter.exp = inter.exp %>% 
              NormalizeData(assay = "RNA", 
                            verbose = F) %>% 
              FindVariableFeatures() %>% 
              ScaleData(features = rownames(.))
  
VariableFeatures(inter.exp@assays$RNA) = inter.exp@assays$RNA@var.features[!(inter.exp@assays$RNA@var.features %in% sex.genes)]

inter.exp@active.assay = "RNA"
inter.exp = inter.exp %>% RunPCA(npcs = 20) %>% 
                          RunUMAP(dims = 1:8) %>%
                          FindNeighbors(dims = 1:8) %>%
                          FindClusters(resolution = c(0.8, seq(0.5, 2, 0.5)))

Idents(inter.exp) = "RNA_snn_res.0.5"
inter.exp = BuildClusterTree(inter.exp, 
                             dims = 1:8, 
                             assay = "RNA", 
                             reorder = T, 
                             features = inter.exp@assays$RNA@var.features)

inter.exp@meta.data$RNA_snn_res.0.5 = factor(inter.exp@meta.data$RNA_snn_res.0.5 , levels = inter.exp@tools$BuildClusterTree$tip.label)

inter.exp@reductions$umap@cell.embeddings[, "UMAP_1"] = inter.exp@reductions$umap@cell.embeddings[, "UMAP_1"] * -1 #Inverting the x axis so we can have more intuitive visualizations, with maturation going from left to right.
```

### Overview Plots
```{r}
DimPlot(inter.exp, label = T, group.by = "RNA_snn_res.0.5") + 
  scale_color_manual(values = met.brewer("VanGogh2", length(levels(inter.exp$RNA_snn_res.0.5)))) + 
  simple + 
  coord_fixed() + 
  labs(title = "Unsupervised Clustering")
``` 

```{r}
DimPlot(inter.exp, label = F, group.by = "age", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_viridis_d(name = "Age", option = "H")+ 
  labs(title = NULL)
```

```{r}
DimPlot(inter.exp, label = F, group.by = "age_group", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  #scale_color_viridis_d()
  scale_color_manual(values=met.brewer("Hokusai3"))+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, label = F, group.by = "sample", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(values=met.brewer("Juarez", 16))+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, raster = F, label = F, group.by = "region", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(name = "Region of Origin", values = region.pal)+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, group.by = "stream_highlight", order = T) + 
   scale_color_manual(values = (region.pal)[3], na.value = "grey85") +
  coord_fixed() +
  NoAxes()+ 
  labs(title = "EC Stream Cells")
```


```{r}
DimPlot(inter.exp, group.by = "dec_highlight", order = T) +
  scale_color_manual(values = (region.pal)[2], na.value = "grey85") +
  coord_fixed() +
  NoAxes()+ 
  labs(title = "Embryonic EC Cells")
```


```{r}
inter.exp@active.assay = "RNA"
(FeaturePlot(inter.exp, c("GFAP", "TNC",  "SOX2", "TOP2A",  "OLIG2", "SOX10", "DCX", "GABRB2", "LHX6", "NR2F2", "PROX1", "PBX3", "SST", "PVALB", "NPY",  "CALB2", "VIP", "RELN", "KIT",  "LAMP5", "TBR1", "SLC17A7", "ADAM28", "PECAM1"), order = T, ncol = 6) &
  simple &
  mysc & 
  coord_fixed() &
  theme(text = element_text(family = "Helvetica")))
#ggsave("inter.exp_overview_featplots.png", height = 14, width = 28)
```

# Saving inter.exp
```{r}
saveRDS(inter.exp, "inter.exp_step1.rds", compress = F)
```

These were the processing steps to generate the basic datasets used in the paper: all.exp and inter.exp

```{r}
sessionInfo()
```


