---
title: "Merging samples"
author: "Marcos Nascimento"
date: "01/02/2024"
output:
  html_document:
    df_print: paged
subtitle: Nascimento + Franjic // SoupX/DoubletFinder/DropletQC Pipeline
---
```{r Setup}
# Loading necessary libraries
library(tidyverse)
library(Seurat)
library(viridis)
library(patchwork)
library(DropletUtils)
library(future)
library(tibble)
library(ComplexHeatmap)
library(GEOquery)

# Additional color scales
library(MetBrewer)
library(NatParksPalettes)

# Parallel processing setup
plan("multicore", workers = 80)
options(future.globals.maxSize = 30 * 1024^3) # 30 GB

# Custom themes and scales for plots
mytheme <- theme_minimal() + 
  theme(axis.line = element_line(),
        axis.ticks = element_line(),
        text = element_text(family = "Helvetica"))

simple <- NoAxes() + NoLegend()
mysc <- scale_color_viridis(option = "A")
region.pal <- c("#5EBFA2", "#F69663", "#731DD8",  "#FB7C7E")

# List of sex-specific genes
sex.genes <- c("TTTY14", "NLGN4Y", "USP9Y", "UTY", "XIST", "RPS4X", "TMSB4X", "TSIX")

#levels for some metadata categories:
age.levels <- c("23GW", "14d", "33d", "54d", "2y", "3y", "13y", "27y", "50y", "51y", "79y")
age.group.levels <- c("Fetal (23GW)", "Infant (14d-54d)", "Toddler (2y-3y)", "Teen (13y)", "Adult (27y-79y)")
region.levels <- c("Germinal Zone", "Embryonic EC", "Migratory Stream", "Postnatal EC")
```

#Creating seurat objects from count matrices 
Count matrices and metadata are downloaded from GEO and saved in a folder named "matrices".
```{r}
load("gsm_samples.RData")

dir.create("matrices")

for (s in 1:nrow(gsm_samples)) {
print(paste("Downloading count matrices. Sample", s, "of", nrow(gsm_samples)))
    
gsm_url <- paste0("https://www.ncbi.nlm.nih.gov/geo/download/?acc=", gsm_samples$gsm[s], "&format=file&file=", gsm_samples$gsm[s], "%5F", gsm_samples$samplenames[s])
barcodes_url <- paste0(gsm_url, "%5Fbarcodes%2Etsv%2Egz")
counts_url <- paste0(gsm_url, "%5Fcounts%2Emtx%2Egz")
genes_url <- paste0(gsm_url, "%5Fgenes%2Etsv%2Egz")
metadata_url <- paste0(gsm_url, "%5Fmetadata%2Ecsv%2Egz")

download.file(barcodes_url, method = "curl", paste0("matrices/", gsm_samples$samplenames[s], "_barcodes.tsv.gz"))
download.file(counts_url, method = "curl", paste0("matrices/", gsm_samples$samplenames[s], "_counts.mtx.gz"))
download.file(genes_url, method = "curl", paste0("matrices/", gsm_samples$samplenames[s], "_genes.tsv.gz"))
download.file(metadata_url, method = "curl",paste0("matrices/", gsm_samples$samplenames[s], "_metadata.csv.gz"))

}

# Extracting all .gz files

gz_files <- dir("matrices", pattern = ".gz")
for (f in gz_files) {
  print(paste("Extracting file", match(f, gz_files), "of", length(gz_files)))
  gunzip(paste0("matrices/", f), destname = paste0("matrices/", gsub(".gz", "", f)))
  unlink(paste0("matrices/", f))
}

download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE199762&format=file&file=GSE199762%5Fsamples%5Ffrom%5FGSE186538%2Etar%2Egz", 
              method = "curl", 
              "matrices/franjic_et_al_samples.tar.gz")
untar(tarfile = "matrices/franjic_et_al_samples.tar.gz", exdir = "matrices/")

#moving files to the right folder
files_to_move <- dir(path = "matrices/samples_from_GSE186538")
for (f in files_to_move) {
 file.rename(paste0("matrices/samples_from_GSE186538/", f) ,paste0("matrices/", f))
}
unlink("matrices/samples_from_GSE186538", recursive = T)
unlink("matrices/franjic_et_al_samples.tar.gz")
```

```{r}
srt_objects <- list()

for (s in c(gsm_samples$samplenames, "hsb231", "hsb237", "hsb628")) {
  cat(paste0("Importing sample ", s, "\n"))
  matrix <- ReadMtx(mtx = paste0("matrices/", s, "_counts.mtx"), 
                    features = paste0("matrices/", s, "_genes.tsv"), feature.column = 1, 
                    cells = paste0("matrices/", s, "_barcodes.tsv"))
  metadata <- read.csv(paste0("matrices/", s, "_metadata.csv"), row.names = 1)
  srt_objects[[s]] <- CreateSeuratObject(counts = matrix, meta.data = metadata)
}
```

Merging samples in a single Seurat object
```{r}
all.exp = merge(srt_objects[[1]], srt_objects[-1])

all.exp@meta.data$region = factor(all.exp@meta.data$region, region.levels)
all.exp@meta.data$age = factor(all.exp@meta.data$age, levels = age.levels)
all.exp@meta.data$age_group = factor(all.exp@meta.data$age_group, levels = age.group.levels)
```

## log Normalization
```{r}
all.exp@active.assay = "RNA"
all.exp = all.exp %>% 
              NormalizeData(assay = "RNA", 
                            verbose = F) %>% 
              FindVariableFeatures() %>% 
              ScaleData()
  
VariableFeatures(all.exp@assays$RNA) = all.exp@assays$RNA@var.features[!(all.exp@assays$RNA@var.features %in% sex.genes)]


all.exp = all.exp %>% RunPCA(npcs = 50)

all.exp = all.exp %>% 
                  RunUMAP(dims = 1:22) %>% 
                  FindNeighbors(dims = 1:22) %>%
                  FindClusters(resolution = c(0.8, seq(0.5, 2, 0.5)))

```


# Overview FeaturePlots
```{r}
FeaturePlot(all.exp, c("GFAP", "HOPX", "TOP2A", "EOMES", "DCX", "TBR1", "SLC17A7", "GAD2", "LHX6", "NR2F2", "PROX1", "CALB2", "LAMP5", "RELN", "VIP", "NPY", "SST", "OLIG2", "SOX10",  "MBP", "FOXJ1", "PECAM1", "PDGFRB", "ADAM28"), order = F, raster = F, ncol = 6) &
  simple &
  mysc & 
  coord_fixed()

ggsave("all.exp_featplots.png", width = 10, height = 6, scale = 2)
```

# Overview Plots
Main Cell Types
```{r}
DimPlot(all.exp, raster = F, label = T, repel = T, group.by = "all.exp_type", shuffle = T) +
  coord_fixed() +
  simple +
  labs(title = "Cell Types")
```

Unsupervised Clustering
```{r}
DimPlot(all.exp, raster = F, label = T, group.by = "RNA_snn_res.1", shuffle = T) + 
  coord_fixed() + 
  simple+
  scale_color_manual(values = met.brewer("VanGogh2", n = 50, override.order = T))+ 
  labs(title = "Unsupervised Clustering")
```

Donor Age
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "age", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_viridis_d(option = "H", name = "Donor Age")+ 
  labs(title = NULL)
```

Donor Age Group
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "age_group", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_manual(values=met.brewer("Hokusai3"), name = "Donor Age Group")+ 
  labs(title = NULL)
```

Sample ID
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "sample", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(values=met.brewer("Juarez", 16), name = "Sample")+ 
  labs(title = NULL)
```

Sample Origin
```{r}
DimPlot(all.exp, raster = F, label = F, group.by = "region", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(name = "Sample Origin", values = region.pal) + 
  labs(title = NULL)
```

EC Stream Cells
```{r}
DimPlot(all.exp, group.by = "stream_highlight", order = T, raster = F) + 
  scale_color_manual(values = (region.pal)[3], na.value = "grey85", labels = c("EC Stream (14d)", "Other cells")) +
  coord_fixed() +
  NoAxes() + 
  labs(title = "EC Stream Cells")
```

Nuclear Fraction
```{r}
FeaturePlot(all.exp, "nuclear_fraction", order = T, raster = F) +
  scale_color_viridis(limits = c(0, 1), name = "Nuclear Fraction") +
  coord_fixed() +
  NoAxes() +
  labs(title = NULL)
```

# Saving all.exp
```{r}
Idents(all.exp) = "all.exp_type"
all.exp <- BuildClusterTree(all.exp, assay = "RNA", reorder = T, features = all.exp@assays$RNA@var.features)

saveRDS(all.exp, file = "all.exp.rds", compress = F)
```

# Subsetting interneurons
```{r}
inter.exp.cells = all.exp@meta.data %>% filter(all.exp_type == "Cortical Interneurons") %>% rownames()
DimPlot(all.exp, label = T, cells.highlight = inter.exp.cells, raster = F) + simple + coord_fixed()

inter.exp = subset(all.exp, cells = inter.exp.cells)

load("reorder_index.Rdata") # load the index for reordering the cells. In addition to using the same seed (42, the default one in Seurat), in order to reproduce the exact same PCA results and UMAP projection, our count matrices should be in the same order as the original ones. For the sake of reproducibility, we will reorder the cells.

inter.exp@assays$RNA@counts <- inter.exp@assays$RNA@counts[ , reorder_index]
inter.exp@assays$RNA@data <- inter.exp@assays$RNA@data[ , reorder_index]
```

## logNormalization
```{r}
inter.exp@active.assay = "RNA"

inter.exp = inter.exp %>% 
              NormalizeData(assay = "RNA", 
                            verbose = F) %>% 
              FindVariableFeatures() %>% 
              ScaleData(features = rownames(.))

VariableFeatures(inter.exp@assays$RNA) = inter.exp@assays$RNA@var.features[!(inter.exp@assays$RNA@var.features %in% sex.genes)]

inter.exp@active.assay = "RNA"
inter.exp = inter.exp %>% RunPCA(npcs = 20) %>% 
                          RunUMAP(dims = 1:8) %>%
                          FindNeighbors(dims = 1:8) %>%
                          FindClusters(resolution = c(0.8, seq(0.5, 2, 0.5)))

Idents(inter.exp) = "RNA_snn_res.0.5"
inter.exp = BuildClusterTree(inter.exp, 
                             dims = 1:8, 
                             assay = "RNA", 
                             reorder = T, 
                             features = inter.exp@assays$RNA@var.features)

inter.exp@meta.data$RNA_snn_res.0.5 = factor(inter.exp@meta.data$RNA_snn_res.0.5 , levels = inter.exp@tools$BuildClusterTree$tip.label)

inter.exp@reductions$umap@cell.embeddings[, "UMAP_1"] = inter.exp@reductions$umap@cell.embeddings[, "UMAP_1"] * -1 #Inverting the x axis so we can have more intuitive visualizations, with maturation going from left to right.
```

### Overview Plots
```{r}
DimPlot(inter.exp, label = T, group.by = "RNA_snn_res.0.5") + 
  scale_color_manual(values = met.brewer("VanGogh2", length(levels(inter.exp$RNA_snn_res.0.5)))) + 
  simple + 
  coord_fixed() + 
  labs(title = "Unsupervised Clustering")
``` 

```{r}
DimPlot(inter.exp, label = F, group.by = "age", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  scale_color_viridis_d(name = "Age", option = "H")+ 
  labs(title = NULL)
```

```{r}
DimPlot(inter.exp, label = F, group.by = "age_group", shuffle = T) + 
  coord_fixed() + 
  NoAxes() + 
  #scale_color_viridis_d()
  scale_color_manual(values=met.brewer("Hokusai3"))+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, label = F, group.by = "sample", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(values=met.brewer("Juarez", 16))+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, raster = F, label = F, group.by = "region", shuffle = T) + 
  coord_fixed() + 
  NoAxes() +
  scale_color_manual(name = "Region of Origin", values = region.pal)+ 
  labs(title = NULL)
```


```{r}
DimPlot(inter.exp, group.by = "stream_highlight", order = T) + 
   scale_color_manual(values = (region.pal)[3], na.value = "grey85") +
  coord_fixed() +
  NoAxes()+ 
  labs(title = "EC Stream Cells")
```


```{r}
DimPlot(inter.exp, group.by = "dec_highlight", order = T) +
  scale_color_manual(values = (region.pal)[2], na.value = "grey85") +
  coord_fixed() +
  NoAxes()+ 
  labs(title = "Embryonic EC Cells")
```


```{r}
inter.exp@active.assay = "RNA"
(FeaturePlot(inter.exp, c("GFAP", "TNC",  "SOX2", "TOP2A",  "OLIG2", "SOX10", "DCX", "GABRB2", "LHX6", "NR2F2", "PROX1", "PBX3", "SST", "PVALB", "NPY",  "CALB2", "VIP", "RELN", "KIT",  "LAMP5", "TBR1", "SLC17A7", "ADAM28", "PECAM1"), order = T, ncol = 6) &
  simple &
  mysc & 
  coord_fixed() &
  theme(text = element_text(family = "Helvetica")))
#ggsave("inter.exp_overview_featplots.png", height = 14, width = 28)
```

# Saving inter.exp
```{r}
saveRDS(inter.exp, "inter.exp_step1.rds", compress = F)
```

These were the processing steps to generate the basic datasets used in the paper: all.exp and inter.exp

```{r}
sessionInfo()
```


