---
title: "hdWGCNA"
author: "Marcos Nascimento"
date: "`r Sys.Date()`"
output: html_notebook
---

# hdWGCNA
This analysis followed the tutorial by Sam Morabito at https://smorabit.github.io/hdWGCNA/articles/basic_tutorial.html 

## Loading Libraries
```{r setup}
# single-cell analysis package
library(Seurat)

# plotting and data science packages
library(tidyverse)
library(cowplot)
library(patchwork)
library(viridis)
library(grDevices)
library(writexl)

# co-expression network analysis packages:
library(WGCNA)
library(hdWGCNA)

# using the cowplot theme for ggplot
theme_set(theme_cowplot())

# set random seed for reproducibility
set.seed(12345)

#Variables to keep plots consistent:
simple = NoAxes() + NoLegend()

mysc = scale_color_viridis(option = "A")
region.pal = c("#5EBFA2", "#F69663", "#731DD8",  "#FB7C7E")

#Custom functions
ggsave <- function(..., bg = "white") ggplot2::ggsave(..., bg = bg)
```


## Setup
```{r}
inter.exp_step3 = readRDS("../4.monocle/inter.exp_step3.rds")

inter.exp_step3@misc$active_wgcna = NULL
inter.exp_step3@misc$inter_fraction_0.05 = NULL

inter.exp_step3@active.assay = "RNA"
inter.exp_step3 = inter.exp_step3 %>% NormalizeData() %>% FindVariableFeatures() 

inter.exp_step3 = inter.exp_step3 %>% SetupForWGCNA(
  gene_select = "fraction", # the gene selection approach
  fraction = 0.05, # fraction of cells that a gene needs to be expressed in order to be included
  wgcna_name = "inter_fraction_0.05" # the name of the hdWGCNA experiment
)

ncol(inter.exp_step3) #20470 cells
```

```{r}
DimPlot(inter.exp_step3, group.by = "all.exp_type") + coord_fixed()
DimPlot(inter.exp_step3, group.by = "sample") + coord_fixed()
```


## Constructing Metacells
```{r}
# construct metacells  in each group
inter.exp_step3 = inter.exp_step3 %>% MetacellsByGroups(
  group.by = c("all.exp_type", "sample"), # specify the columns in seurat_obj@meta.data to group by
  k = 20, # nearest-neighbors parameter
  max_shared = 10, # maximum number of shared cells between two metacells
  ident.group = "all.exp_type", # set the Idents of the metacell seurat object
)

# normalize metacell expression matrix:
inter.exp_step3 = NormalizeMetacells(inter.exp_step3)
```

## Process the Metacell Seurat Object
Optional step, but may be useful since metacells are pseudobulk cells with way less sparsity and noise than the original nuclei
```{r}
metacell_objnew <- GetMetacellObject(inter.exp_step3)

inter.exp_step3 <- NormalizeMetacells(inter.exp_step3)
inter.exp_step3 <- ScaleMetacells(inter.exp_step3, features=VariableFeatures(inter.exp_step3))
inter.exp_step3 <- RunPCAMetacells(inter.exp_step3, features=VariableFeatures(inter.exp_step3))
inter.exp_step3 <- RunUMAPMetacells(inter.exp_step3, reduction='pca', dims=1:20)


p1 <- DimPlotMetacells(inter.exp_step3, group.by='all.exp_type') + umap_theme() + ggtitle("Cell Type")
p2 <- DimPlotMetacells(inter.exp_step3, group.by='sample') + umap_theme() + ggtitle("Sample")

p1 | p2
```

## Co-expression network analysis
```{r}
inter.exp_step3 = inter.exp_step3 %>%  SetDatExpr(
  group_name = c("Cortical Interneurons"), # the name of the group of interest in the group.by column
  group.by='all.exp_type', # the metadata column containing the cell type info. This same column should have also been used in MetacellsByGroups
  assay = 'RNA', # using RNA assay
  slot = 'data' # using normalized data
)
```

## Select soft-power threshold
```{r}
# Test different soft powers:
inter.exp_step3 <- TestSoftPowers(
  inter.exp_step3,
  networkType = 'signed' # you can also use "unsigned" or "signed hybrid"
)

# plot the results:
plot_list <- PlotSoftPowers(inter.exp_step3)

# assemble with patchwork
wrap_plots(plot_list, ncol=2)
```

## Construct co-expression network
```{r}
# construct co-expression network:
inter.exp_step3 <- ConstructNetwork(
  inter.exp_step3, soft_power=3, #select the appropriate soft power determined in the previous code chunk
  setDatExpr=FALSE,
  tom_name = 'INTER', # name of the topoligical overlap matrix written to disk
  overwrite_tom = T 
)

PlotDendrogram(inter.exp_step3, main='topo_1 hdWGCNA Dendrogram')
```

## Topological Overlap Matrix (TOM)
```{r}
TOM <- GetTOM(inter.exp_step3)
```

## Compute module eigengenes
```{r}
# compute all MEs in the full single-cell dataset
inter.exp_step3 <- ModuleEigengenes(
 inter.exp_step3,
 exclude_grey = T
)
```
## Compute module connectivity
```{r}
# compute eigengene-based connectivity (kME):
inter.exp_step3 <- ModuleConnectivity(
  inter.exp_step3,
  group.by = 'all.exp_type', 
  group_name = 'Cortical Interneurons'
)

# rename the modules
inter.exp_step3 <- ResetModuleNames(
  inter.exp_step3,
  new_name = "INH-M"
)

# plot genes ranked by kME for each module
p <- PlotKMEs(inter.exp_step3, ncol=7, text_size = 4)

for(m in 1:5) {
  wrap_plots(p[[m]])
  ggsave(paste0("plots/module-connectivity_M", m, ".png"), width = 4, height = 4, scale = 1)
}
```
## Getting the module assignment table
```{r}
# get the module assignment table:
modules <- GetModules(inter.exp_step3)

# show the first rows:
head(modules)

# get hub genes
hub_df <- GetHubGenes(inter.exp_step3, n_hubs = 200)

#Exporting as an excel file:
hub_df = hub_df %>% mutate(paper_module = module)
hub_df = hub_df %>% arrange(paper_module, -kME) %>% select(paper_module, gene_name, kME)
colnames(hub_df) = c("module", "gene", "kME")
#write_xlsx(hub_df, path = "wgcna_modules_genes.xlsx")
```

#Saving inter.exp_step4
```{r}
inter.exp_step4 = inter.exp_step3
saveRDS(inter.exp_step4, "inter.exp_step4.rds", compress = F)
```


## Ploting eigengenes
```{r}
# make a featureplot of hMEs for each module
me_plot_list <- ModuleFeaturePlot(
  inter.exp_step4,
  features='MEs', # plot the hMEs
  order=TRUE # order so the points with highest hMEs are on top
)

# stitch together with patchwork
me_plot_list #Better

for(m in 1:5) {
  me_plot_list[[m]] + scale_color_gradient2(low = "grey", mid = "white", high = "red")
  ggsave(paste0("plots/eigengene_featureplots_M", m, ".png"), width = 4, height = 4, scale = 2)
}
```

##Violin Plots
```{r}
wgcnadata = merge(inter.exp_step4@meta.data, inter.exp_step4@misc[["inter_fraction_0.05"]][["MEs"]], by = 0) %>% column_to_rownames(var = "Row.names")
wgcnadata = wgcnadata %>% mutate(`scaled_INH-M1` = scale.default(`INH-M1`))
wgcnadata = wgcnadata %>% mutate(`scaled_INH-M2` = scale.default(`INH-M2`))
wgcnadata = wgcnadata %>% mutate(`scaled_INH-M3` = scale.default(`INH-M3`))
wgcnadata = wgcnadata %>% mutate(`scaled_INH-M4` = scale.default(`INH-M4`))
wgcnadata = wgcnadata %>% mutate(`scaled_INH-M5` = scale.default(`INH-M5`))

EC_14d_cells = wgcnadata %>% filter(age == "14d" & region == "Postnatal EC") %>% rownames
stream_14d_cells = wgcnadata %>% filter(age == "14d" & region == "Migratory Stream") %>% rownames
ge_cells = wgcnadata %>% filter(age == "23GW" & region == "Germinal Zone") %>% rownames
embryonic_ec_cells = wgcnadata %>% filter(age == "23GW" & region == "Embryonic EC") %>% rownames

wgcnadata = wgcnadata %>%  mutate(plot_ages = paste(age, "EC"))
wgcnadata[EC_14d_cells, "plot_ages"] = "14d EC"
wgcnadata[stream_14d_cells, "plot_ages"] = "14d Migratory Stream"
wgcnadata[ge_cells, "plot_ages"] = "23GW Germinal Zone"
wgcnadata[embryonic_ec_cells, "plot_ages"] = "23GW EC"

wgcnadata$plot_ages = factor(wgcnadata$plot_ages, levels = c("23GW Germinal Zone", 
                                                             "23GW EC", "14d Migratory Stream", 
                                                             "14d EC", "33d EC", 
                                                             "54d EC", 
                                                             "2y EC", 
                                                             "3y EC", 
                                                             "13y EC", 
                                                             "27y EC", 
                                                             "50y EC", 
                                                             "51y EC", 
                                                             "79y EC"))


INH_mods = paste0("INH-M", seq(1:5))

#plotting vertical versions of the violin plots:
for( i in 1: length(INH_mods)) {
wgcnadata %>% 
  #filter(all.exp_type == "Cortical Interneurons") %>% 
  ggplot(aes(get(INH_mods[i]), fct_rev(plot_ages), fill = region)) + 
  geom_violin() +
  scale_fill_manual(name = "Region", 
                    values = region.pal) +
  theme(axis.text.x = element_blank()) + 
  scale_y_discrete() +
  labs(y = NULL, 
       x = NULL) + 
  theme_minimal() + 
  theme(axis.line = element_line(), 
        panel.grid.major.y = element_blank(), 
        legend.position = "none")
  
ggsave(paste0("plots/module", i, "_violinplot.by.ages_vertical.png"), width = 4, height = 8, scale = 1) 
}
```


```{r}
wgcna.list = list()
for(m in INH_mods) {
  wgcna.list[[m]] = merge(select(wgcnadata, age, region, pseudotime), select(wgcnadata,m), by = 0)
  colnames(wgcna.list[[m]])[5] = "module"
}

myplots = list()
  for(m in INH_mods) {
     myplots[[m]] = wgcna.list[[m]] %>% 
                      ggplot(aes(x = pseudotime, y = module)) + 
                      geom_jitter(aes(col = age), width = 0.5, alpha = 0.5, size = 0.5) + 
                      scale_color_viridis_d(option = "H") + 
                      geom_smooth(col = "red") + 
                      theme_minimal() + 
                      scale_x_continuous(expand = c(0,0)) +
                      labs(x = "Pseudotime", y = "Module eigengene") + 
                      theme(legend.position = "none")
     wrap_plots(myplots[[m]])
     ggsave(paste0("plots/wgcna_modules_pseudotime_", m, ".pdf"), width = 4, height = 4, scale = 0.8)
  }

wrap_plots(myplots, ncol = 7)
ggsave("wgcna_modules_pseudotime.png", width = 16, height = 2.5, scale = 2)


myplots = list()
  for(m in INH_mods) {
     myplots[[m]] = wgcna.list[[m]] %>% 
                      ggplot(aes(x = pseudotime, y = module)) + 
                      geom_jitter(aes(col = region), width = 0.5, alpha = 0.5, size = 0.5) + 
                      scale_color_manual(values = region.pal) + 
                      geom_smooth(col = "red") + 
                      theme_minimal() + 
                      scale_x_continuous(expand = c(0,0)) +
                      labs(x = "Pseudotime", y = "Module eigengene") + 
                      theme(legend.position = "none")
     wrap_plots(myplots[[m]])
     ggsave(paste0("plots/wgcna_modules_pseudotime_byregion_", m, ".pdf"), width = 4, height = 4, scale = 0.8)
  }

wrap_plots(myplots, ncol = 7)
ggsave("wgcna_modules_pseudotime_byregion.png", width = 16, height = 2.5, scale = 2)
```

# Module Trait Correlation
## Calculating
```{r}
# convert age to numeric
inter.exp_step4$age %>% unique()
inter.exp_step4@meta.data = inter.exp_step4@meta.data  %>% mutate(age_c = ifelse(age == "23GW", 0.4, 
                                          ifelse(age == "14d", 0.78,
                                          ifelse(age == "33d", 0.83,
                                          ifelse(age == "54d", 0.89,
                                          ifelse(age == "2y", 2.74,       
                                          ifelse(age == "3y", 3.74,        
                                          ifelse(age == "13y", 13.74,
                                          ifelse(age == "27y", 27.74, 
                                          ifelse(age == "50y", 50.74,        
                                          ifelse(age == "51y", 51.74,        
                                          ifelse(age == "79y", 79.74,
                                                 "unknown"))))))))))))
                                          

inter.exp_step4$age_c <- as.numeric(inter.exp_step4$age_c)

# list of traits to correlate
cur_traits <- c('age_c', 'pseudotime', 'nCount_RNA', 'percent.mt', "nuclear_fraction")

inter.exp_step4 <-ModuleTraitCorrelation(
  inter.exp_step4,
  traits = cur_traits
)
```
## Inspecting
```{r}
# get the mt-correlation results
mt_cor <- GetModuleTraitCorrelation(inter.exp_step4)
mt_cor$cor$all_cells
mt_cor$fdr$all_cells
```


## Plotting
```{r}
ggplot(mt_cor$cor$all_cells %>% as.data.frame() %>% rownames_to_column("variable") %>% pivot_longer(cols = 2:6, names_to = "module") %>% mutate(label = sprintf("%.2f", value)), aes(module, variable, fill = value, label = label)) + 
  geom_tile() + 
  geom_text() + 
  scale_fill_distiller(type = "div", palette = 5, limits = c(-1, 1), name = "Pearson coefficient") + 
  theme_minimal() + 
  theme(panel.grid = element_blank()) +
  labs(x = NULL, y = NULL) + 
  scale_y_discrete(labels = c("Age", "UMI count", "Nuclear Fraction", "Mitochondrial Fraction", "Pseudotime"))
```

